\bf Visual Inference for a Social Network Model

### 2019-03-05 ASApaper/paper-child.Rnw
done with big revision

\noindent \textbf{Objective Function:} After actor $i$ has been selected by the rate function to change, it randomly picks one of its current ties, $x_{ij}$, to change. The probability that actor $i$ changes its current tie to actor $j$ is determined by the \textit{objective function} of the model and a random error term $U$ with log-Weibull distribution \citep{modelsSnijders}.  %\hh{XXX can you paraphrase the purpose of U in half a sentence?}
%The random component is assumed to have log-Weibull distribution with location parameter $\mu = 0$ and scale parameter $\sigma = 0$ (see \citet{modelsSnijders}), so the probability density function of $U$ is
%\begin{equation}\label{eq:logweibull}
%g(u) = \exp\{-(u + \exp\{-u\})\}.
%\end{equation}
%The objective function, $f_i$, is the driving force of a SAOM. 
Actor $i$ aims to maximize the objective function $f_i$ given the current state of the network, $x$ and the node-level covariates, $\mathbf{Z}$. This function is defined as:% which has the form
\begin{equation}\label{eq:objective}
f_i(x, \boldsymbol{\beta}, \mathbf{Z}) = \sum_{k = 1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)$ are additional model parameters, each associated with some statistics, $s_{i1}(x, \mathbf{Z})$, \dots, s_{iK}(x, \mathbf{Z})$, calculated for actor $i$ at the current network state $x$. The statistics can be very simple, like outdegree of the node, $s_i(x) = \sum_{i\neq j} x_{ij}$, or more complicated, like the {\it transitive triplets jumping to different covariate} statistic, $s_i(x, \mathbf{Z}) = \sum_{i \neq j \neq h} x_{ij}x_{ih}x_{hj} \cdot \mathbb{I}(z_i = z_h \neq z_j)$. %, plus many more. At last count, in the software we use to fit these models to network data, 
Version 1.2-3 of \texttt{RSiena} \citep{RSiena}, the software used to fit the models here, provides over 80 possible effects that can be included in the objective function. We discuss these and other statistics in more detail in Section~\ref{sec:models}.

The objective function $f_{i}(x, \boldsymbol{\beta}, \mathbf{Z})$ and  random component $U$ combine to form the \textit{transition probability}, $p_{ij}$ of the network changing from its current state $x$ to the state $x(i\leadstoj)$, which is identical to $x$ except for $x_{ij}$: $x_{ij}(i \leadsto j) = 1 - x_{ij}$. The transition probability is
\begin{equation}\label{eq:transprob}
p_{ij} = \dfrac{\exp\{f_i(x(i \leadsto j), \boldsymbol{\beta}, \mathbf{Z})\}}{\sum_{h} \exp\{f_i(x(i \leadsto h), \boldsymbol{\beta}, \mathbf{Z})\}}
\end{equation}
This probability dictates which edge change is made by the node $i$. Node $i$ can also choose to \textit{not} change at all, which is most likely when the value of Equation~\ref{eq:objective} is larger for $x$ than for any $x(i \leadsto j)$.

@@

<<senateAll, fig.cap="The US senate collaboration network observed at four time points. Color represents party, shape represents gender, and size represents number of bills authored in a session. The Frucherman-Reingold layout is shown \citep{fr}. Drawn with the \\texttt{geomnet} R package. \\citep{geomnet, rstats}">>=
seobama <- read_csv("data/senateobamapres_gsw_25.csv")
to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar", "Elizabeth Warren")
seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Elizabeth Warren",
                           "Harry Reid", "Hillary Clinton", "Jim Webb", "John McCain", 
                           "Joe Biden","Lindsey Graham", "Marco Rubio", "Mitch McConnell", 
                           "Rand Paul", "Ted Cruz")
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
seobama2 <- seobama %>% filter(senate != 111) %>% bind_rows(se111noclint)
mynames <- data.frame(do.call(rbind, stringr::str_split(seobama2$label, " ", n = 2) ))
names(mynames) <- c("first", "last")
seobama2$label <- mynames$last

@@

\begin{table}
\scalebox{0.8}{
\begin{tabular}{c|p{2cm}|p{2.1cm}|c|l|p{1.75cm}|p{2cm}}
$\beta_k$ & {\bf Effect name} & {\bf Interaction Variable} & {\bf Formula} & {\bf Picture} & {\bf Initial estimate} & {\bf Wald $p$-value} \\
\hline 
$\beta_1$ & density & -- & $s_{i1}(x) = \sum_{j} x_{ij}$ & \includegraphics[width=.6in]{img/density.png} & 2.204 & NA \\ 
$\beta_2$ & reciprocity & -- & $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$ & \includegraphics[width=.6in]{img/recip.png} & -4.903 & NA \\
$\beta_3$ & jumping transitive triplet & party & $s_{i3}(x, \mathbf{p}) = \sum_{j\neq h} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(p_i = p_h \neq p_j)$ & \includegraphics[width=.6in]{img/jttp.png} & -5.884 & $<0.0001$\\
$\beta_4$ & jumping transitive triplet & sex & $s_{i4}(x, \mathbf{s}) = \sum_{j\neq h} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(s_i = s_h \neq s_j)$ & \includegraphics[width=.6in]{img/jtts.png}  & 3.335 & 0.0002 \\
$\beta_5$ & similarity transitive triplet & bills & $s_{i5}(x, \mathbf{b}) = \sum_{j} x_{ij}x_{ih}x_{hj}\cdot (sim^b_{ij} - \overline{sim}^b)^*$ & \includegraphics[width=1in]{img/simttb.png} & 9.821 & 0.0128 \\
$\beta_6$ & same transtive triplet & party & $s_{i6}(x, \mathbf{p}) =\sum_{j} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(p_i = p_j)$ & \includegraphics[width=1in]{img/samettp.png}  & 1.306 & 0.0642 %\\
%$\beta_7$ & same & party & $s_{i7}(x, \mathbf{p}) = \sum_j x_{ij}\mathbb{I}(p_i = p_j)$ & \includegraphics[width=.6in]{img/samep.png}  & 0.363 & 0.0074
\end{tabular}
}
\caption{\label{tab:effects} The effects we used in the CTMC models fit to the senate data. In the picture which represents each effect, the dotted tie is encouraged to form if the estimate is positive, and discouraged to form if the estimate is negative.  $^*$ - $sim^b_{ij} is defined in Equation~\ref{eq:similar} and $\overline{sim}^b = \frac{1}{n(n-1)}\sum_{i\neq j} sim^b_{ij}$ is the average bill similarity score between two senators.}
\end{table}

@@

Because the size of the lineups is small, the probability of picking the data by chance is high, $\frac{1}{6}$, but if \textit{many} independent viewers pick out the data from the nulls, the evidence in favor of the alternative hypothesis becomes stronger.

@@

<<sigtesting, results='asis'>>=
turk22_sig <- turk22 %>% 
  mutate(type2 = ifelse(sign == 0, sign(initialEst), sign),
         type2 = ifelse(type == "one", type2, -1)) %>% 
  filter(param_value == 1)
summ_sig <- turk22_sig %>% group_by(pic_id, test_param) %>% 
  summarize(npick = sum(datapick), n = n(), 
            pv = map2_dbl(npick, n, pV, m=6, scenario=3))
names(summ_sig) <- c("Lineup ID", "parameter", "# Alt. Model Picks", "Total Views", "p-value")
summ_sig$parameter <- as.factor(summ_sig$parameter)
levels(summ_sig$parameter) <- c("$ \\beta_3 $", "$ \\beta_4$ ")
xtable::print.xtable(xtable::xtable(summ_sig, caption = "Experiment results for the two parameters for which we performed significance tests. There were three lineups for each parameter, so there are three results for each plot.", label = "tab:sigtesting", digits = c(0,0,0,0,0,5)), include.rownames = FALSE, sanitize.text.function=function(x){x})
@

The $p$-values were calculated using the \texttt{vinference} package by \citet{vinference}. This package contains methods to calculate \textit{Visual distributions} for lineup experiment data.  The distribution depends on the number of evaluations of a plot, $L$, the size of the lineup, $P$, and the lineup scenario, which here is that each lineup containing the same data and the same set of null plots is shown to $L$ independent observers. The visual inference family of distributions is similar to the binomial distribution, but takes the dependency among the $P$ plots in a single lineup shown to multiple viewers into account. %Using these $p$-values, all but one lineup results in a rejection of the null hypothesis at Type-I error rate of $\alpha = 0.05$. 
We see in Table~\ref{tab:sigtesting} that the $p$-values for visual inference here are highly variable.

@@

Goodness-of-fit testing for network models is notoriously difficult. Most network models, other than the most simple, lack the necessary asymptotics for developing goodness-of-fit methods \citep{goldenberg09}. Some simulation-based methods have been developed using what \citeauthor{RSienamanual} call \textit{auxiliary statistics} such as the indegree or outdegree distribution on the nodes. In \texttt{RSiena}, the \texttt{sienaGOF} function performs goodness-of-fit testing as follows:
\begin{enumerate}
\item Auxiliary statistics, such as the cumulative outdegree counts on the nodes, are computed on the observed data ($\mathbf{u}_d$) and on $N$ observations simulated from the model ($\mathbf{u}_1 \dots \mathbf{u}_N$). 
\item The mean $\overline{\mathbf{u}}$ and covariance matrix $\mathbf{S}$ are computed from the $N$ simulations, and the Mahalanobis distance, $d_M(\mathbf{u}_d)$ from the observed statistics to the distribution of the simulated statistics is computed:
\begin{equation}\label{eq:mahal}
d_M(\mathbf{u}_d) = \sqrt{(\mathbf{u}_d - \overline{\mathbf{u}})' \mathbf{S}^{-1} (\mathbf{u}_d - \overline{\mathbf{u}})}
\end{equation}
\item The Mahalanobis distance for each of the $N$ simulations is calculated and $d_M(\mathbf{u}_d)$ is compared to this distribution of distances.
\item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than $d_M(\mathbf{u}_d)$.% A model is thus considered a good fit to the data if $p$ is large.
\end{enumerate}
<<gofsiena, fig.height=2.5, fig.cap="An example of what a goodness-of-fit plot from \\texttt{RSiena} looks like. The overlaid boxplots and violin plots show the distribution of each of the outdegree count values on the simulated networks, and the red points and lines are the observed data values.", cache = TRUE>>=
load("data/ansnullpaper.rda")
library(lattice)
gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
sims1 <- data.frame(gof1$`Period 1`$Simulations)
dat1 <- gof1$`Period 1`$Observations
dat1 %>% data.frame() %>% gather(outdegree, val) %>%
  mutate(outdegree = parse_number(outdegree)-1) -> dat1
p1 <- sims1 %>%
  mutate(sim = row_number()) %>%
  gather(outdegree, val, X1:X9) %>%
  mutate(outdegree = parse_number(outdegree) - 1) %>%
  filter(outdegree <= 6) %>%
  ggplot(aes(x = outdegree, y = val)) + 
  geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
  geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
  geom_point(data = dat1, color = 'red') + 
  geom_line(data= dat1, color = 'red') + 
  geom_text(data = dat1, aes(label = val), hjust = -.5) + 
  labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
       title = "Goodness-of-Fit: Outdegree distribution period 1") + 
  ThemeNoNet
p1
@
A plot comparing the data to the simulations is also considered, and a similar plot is shown in Figure~\ref{fig:gofsiena} for the outdegree distribution of small data set, shown in the points and connected lines, with the simulated values of $\mathbf{u}_d$ shown in boxplots and overlaid violin plots.
The \texttt{RSiena} software also provides a Rao score-type test for goodness-of-fit for assessing one or more parameters, the test statistic of which is compared to a Chi-square distribution with $H$ degrees of freedom, where $H$ has the same definition as in Section~\ref{sec:sigtest}. For full detail on the score-type test, see \citet{scoretest}. These methods are both restriced: the \texttt{sienaGOF} method only considers \textit{one} measure on the data and simulations from the model, while the score-type tests only consider \textit{subsets} of parameters, ``nuisance parameters" in \citet{scoretest}, not the entire set of parameters. By using visual inference instead of more traditional statistical methods, we hope to perform a more \textit{holistic} goodness-of-fit test.

@@

<<predictglmm, fig.height=6, fig.cap='Predictions from our generalized linear mixed effects model given in Equation~\ref{eq:glmm}. The lines show the expected probability of detecting the alternative data in a lineup of size 6 for new observers of new lineups is plotted on the $y$-axis, and the size of the parameter of interest is on the $x$-axis. The proportions detected by our Turk participants for each lineup group are shown by the points, with the probability of picking out the data plot at random shown by a horizontal line at 1/6. The lineup marked as ``outlier" was removed from modeling. The panel for the reciprocity parameter, $\\beta_2$ is also presented in Figure~\\ref{fig:recipzoom} in more detail.'>>=
newdata <- read_csv("data/newdata_pred_glmm.csv")
newdata$param <- newdata$test_param
newdata$param <- as.factor(newdata$test_param)
levels(newdata$param) <-  c(1,3,4,2,6,5) 
newdata$param <- as.integer(as.character(newdata$param))
plotdata <- turk22 %>% mutate(type2 = ifelse(sign == 0, sign(initialEst), sign),
                       type2 = ifelse(type == "one", type2, -1)) %>% 
  group_by(pic_id, size, test_param, sign, type2,alt_model) %>% summarize(
    datapick = mean(datapick)
)
plotdata$param <- plotdata$test_param
plotdata$param <- as.factor(plotdata$param)
levels(plotdata$param) <- c(1,3,4,2,6,5) 
plotdata$param <- as.integer(as.character(plotdata$param))
labdat <- data_frame(test_param = "jttp", param = 3, type2 = -1, x = -5, y = .9, label = "jttp_neg_hard_2")

ggplot() + 
  geom_line(data = newdata, aes(x = size, y = predictfinal, color = as.factor(type2))) + 
  geom_point(data = plotdata, aes(x = size, y = datapick, color = as.factor(type2))) + 
  #geom_text(data = plotdata, aes(x = size, y = datapick, color = as.factor(type2), label = pic_id)) +
  geom_text(data = labdat, aes(x = x, y=y, label = label, color = as.factor(type2)),
            show.legend = F) + 
  geom_hline(yintercept = 1/6, linetype='dashed') +
  scale_color_brewer(palette = "Dark2", name = "Lineup Type") + 
  facet_wrap(~param, scales = 'free_x', 
             labeller = label_bquote(beta [.(param)])) + 
  labs(x = "Parameter value", y = "Expected probability of detection in a lineup of size 6") + 
  ThemeNoNet + 
  theme(legend.position = 'bottom')

@@

These results do not come without limitations. In VI, the null plots are supposed to play the role of good representatives of the null model. Here, the number of null plots is reduced to five, which increases the variability seen in a single lineup dramatically, and can unfortunately lead to very different conclusions for the same lineup scenario, as we have seen. Furthermore, these results do not generalize to all CTMC models or to one particular subset of CTMC models. The lineups shown are made for only one set of data, and it is not clear that the visual power results will transfer to other situations with different number of actors, different edge densities, or different layout algorithm of the node-link diagram. We can make some generalizations about what participants are picking up on in the lineups based on their feedback and previous research, but we cannot apply our hierarchical model directly to lineups constructed for new data or new models or parameters.

@@

We these methods will be applied to different types of network data and different types of network models. But given the limitations of the node-link visualization, the cognitive load of looking at a lineup is very high for the average observer. We would therefore like to explore larger datasets, different layout algorithms, and different ways of visualizing network data, such adjacency matrix visualizations, using visual inference to see if similar patterns emerge.

>>>>

\noindent \textbf{Objective Function:} After actor $i$ has been selected by the rate function to change, it randomly picks one of its current ties, $x_{ij}$, to change. The probability that actor $i$ changes its current tie to actor $j$ is determined by the \textit{objective function} of the model and a random error term $U$ with log-Weibull distribution \citep{modelsSnijders}.  %\hh{XXX can you paraphrase the purpose of U in half a sentence?}
%The random component is assumed to have log-Weibull distribution with location parameter $\mu = 0$ and scale parameter $\sigma = 0$ (see \citet{modelsSnijders}), so the probability density function of $U$ is
%\begin{equation}\label{eq:logweibull}
%g(u) = \exp\{-(u + \exp\{-u\})\}.
%\end{equation}
%The objective function, $f_i$, is the driving force of a SAOM. 
Actor $i$ aims to maximize the objective function $f_i$ given the current state of the network, $x$ and the node-level covariates, $\mathbf{Z}$. This function is defined as:% which has the form
\begin{equation}\label{eq:objective}
f_i(x, \boldsymbol{\beta}, \mathbf{Z}) = \sum_{k = 1}^K \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}
where $\boldsymbol{\beta} = (\beta_1, \dots, \beta_K)$ are additional model parameters, each associated with some statistics, $s_{i1}(x, \mathbf{Z}), \dots, s_{iK}(x, \mathbf{Z})$, calculated for actor $i$ at the current network state $x$. The statistics can be very simple, like outdegree of the node, $s_i(x) = \sum_{i\neq j} x_{ij}$, or more complicated, like the {\it transitive triplets jumping to different covariate} statistic, $s_i(x, \mathbf{Z}) = \sum_{i \neq j \neq h} x_{ij}x_{ih}x_{hj} \cdot \mathbb{I}(z_i = z_h \neq z_j)$. %, plus many more. At last count, in the software we use to fit these models to network data, 
Version 1.2-3 of \texttt{RSiena} \citep{RSiena}, the software used to fit the models here, provides over 80 possible effects that can be included in the objective function. We discuss these and other statistics in more detail in Section~\ref{sec:models}.

The objective function $f_{i}(x, \boldsymbol{\beta}, \mathbf{Z})$ and  random component $U$ combine to form the \textit{transition probability}, $p_{ij}$ of the network changing from its current state $x$ to the state $x(i\leadsto j)$, which is identical to $x$ except for $x_{ij}$: $x_{ij}(i \leadsto j) = 1 - x_{ij}$. The transition probability is
\begin{equation}\label{eq:transprob}
p_{ij} = \dfrac{\exp\{f_i(x(i \leadsto j), \boldsymbol{\beta}, \mathbf{Z})\}}{\sum_{h} \exp\{f_i(x(i \leadsto h), \boldsymbol{\beta}, \mathbf{Z})\}}
\end{equation}
This probability dictates which edge change is made by the node $i$. Node $i$ can also choose to \textit{not} change at all, which is most likely when the value of Equation~\ref{eq:objective} is larger for $x$ than for any $x(i \leadsto j)$.

@@

<<senateAll, fig.cap="The US senate collaboration network observed at four time points. Color represents party, shape represents gender, and size represents number of bills authored in a session. The Fruchterman-Reingold layout is shown \\citep{fr}. Drawn with the \\texttt{geomnet} R package. \\citep{geomnet, rstats}">>=
seobama <- read_csv("data/senateobamapres_gsw_25.csv")
to_label <- c("Joseph R. Biden Jr.", "John S. McCain", "Ted Cruz","Marco Rubio",
  "Lindsey O. Graham","Rand Paul", "Bernard Sanders", "Jim Webb", "Mitch McConnell",
  "Harry M. Reid", "Hillary Rodham Clinton", "Amy Jean Klobuchar", "Elizabeth Warren")
seobama$label <- as.factor(ifelse(seobama$source %in% to_label, seobama$source, ""))
levels(seobama$label) <- c("", "Amy Klobuchar", "Bernie Sanders", "Elizabeth Warren",
                           "Harry Reid", "Hillary Clinton", "Jim Webb", "John McCain", 
                           "Joe Biden","Lindsey Graham", "Marco Rubio", "Mitch McConnell", 
                           "Rand Paul", "Ted Cruz")
se111clint <- filter(seobama, senate == 111)
se111noclint <- filter(seobama, senate == 111)
se111noclint$target[which(se111noclint$target == "Hillary Rodham Clinton")] <- NA
seobama2 <- seobama %>% filter(senate != 111) %>% bind_rows(se111noclint)
mynames <- data.frame(do.call(rbind, stringr::str_split(seobama2$label, " ", n = 2) ))
names(mynames) <- c("first", "last")
seobama2$label <- mynames$last

@@

\begin{table}
\scalebox{0.8}{
\begin{tabular}{c|p{2cm}|p{2.1cm}|c|l|p{1.75cm}|p{2cm}}
$\beta_k$ & {\bf Effect name} & {\bf Interaction Variable} & {\bf Formula} & {\bf Picture} & {\bf Initial estimate} & {\bf Wald $p$-value} \\
\hline 
$\beta_1$ & density & -- & $s_{i1}(x) = \sum_{j} x_{ij}$ & \includegraphics[width=.6in]{img/density.png} & 2.204 & NA \\ 
$\beta_2$ & reciprocity & -- & $s_{i2}(x) = \sum_{j} x_{ij}x_{ji}$ & \includegraphics[width=.6in]{img/recip.png} & -4.903 & NA \\
$\beta_3$ & jumping transitive triplet & party & $s_{i3}(x, \mathbf{p}) = \sum_{j\neq h} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(p_i = p_h \neq p_j)$ & \includegraphics[width=.6in]{img/jttp.png} & -5.884 & $<0.0001$\\
$\beta_4$ & jumping transitive triplet & sex & $s_{i4}(x, \mathbf{s}) = \sum_{j\neq h} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(s_i = s_h \neq s_j)$ & \includegraphics[width=.6in]{img/jtts.png}  & 3.335 & 0.0002 \\
$\beta_5$ & similarity transitive triplet & bills & $s_{i5}(x, \mathbf{b}) = \sum_{j} x_{ij}x_{ih}x_{hj}\cdot (sim^b_{ij} - \overline{sim}^b)^*$ & \includegraphics[width=1in]{img/simttb.png} & 9.821 & 0.0128 \\
$\beta_6$ & same transtive triplet & party & $s_{i6}(x, \mathbf{p}) =\sum_{j} x_{ij}x_{ih}x_{hj}\cdot \mathbb{I}(p_i = p_j)$ & \includegraphics[width=1in]{img/samettp.png}  & 1.306 & 0.0642 %\\
%$\beta_7$ & same & party & $s_{i7}(x, \mathbf{p}) = \sum_j x_{ij}\mathbb{I}(p_i = p_j)$ & \includegraphics[width=.6in]{img/samep.png}  & 0.363 & 0.0074
\end{tabular}
}
\caption{\label{tab:effects} The effects we used in the CTMC models fit to the senate data. In the picture which represents each effect, the dotted tie is encouraged to form if the estimate is positive, and discouraged to form if the estimate is negative.  $^*$ - $sim^b_{ij}$ is defined in Equation~\ref{eq:similar} and $\overline{sim}^b = \frac{1}{n(n-1)}\sum_{i\neq j} sim^b_{ij}$ is the average bill similarity score between two senators.}
\end{table}

@@

In this section, we present the results from our experiment and compare them to traditional statistical tests where applicable.  Because each lineup shown to participants has only six plots, the probability of picking the data by chance is high, $\frac{1}{6}$, but if \textit{many} independent viewers pick out the data from the nulls, the evidence in favor of the alternative hypothesis becomes stronger. The $p$-values from the lineups were calculated using the \texttt{vinference} package by \citet{vinference}. This package contains methods to calculate \textit{Visual distributions} for lineup experiment data.  The distribution depends on the number of evaluations of a plot, $L$, the size of the lineup, $P$, and the lineup scenario, which here is that each lineup containing the same data and the same set of null plots is shown to $L$ independent observers. The visual inference family of distributions is similar to the binomial distribution, but takes the dependency among the $P$ plots in a single lineup shown to multiple viewers into account.

@@

<<sigtesting, results='asis'>>=
turk22_sig <- turk22 %>% 
  mutate(type2 = ifelse(sign == 0, sign(initialEst), sign),
         type2 = ifelse(type == "one", type2, -1)) %>% 
  filter(param_value == 1)
summ_sig <- turk22_sig %>% group_by(pic_id, test_param) %>% 
  summarize(npick = sum(datapick), n = n(), 
            pv = map2_dbl(npick, n, pV, m=6, scenario=3))
names(summ_sig) <- c("Lineup ID", "parameter", "\\# Alt. Model Picks", "Total Views", "p-value")
summ_sig$parameter <- as.factor(summ_sig$parameter)
levels(summ_sig$parameter) <- c("$ \\beta_3 $", "$ \\beta_4$ ")
xtable::print.xtable(xtable::xtable(summ_sig, caption = "Experiment results for the two parameters for which we performed significance tests. There were three lineups for each parameter, so there are three results for each plot.", label = "tab:sigtesting", digits = c(0,0,0,0,0,5)), include.rownames = FALSE, sanitize.text.function=function(x){x})
@

 %Using these $p$-values, all but one lineup results in a rejection of the null hypothesis at Type-I error rate of $\alpha = 0.05$. 
We see in Table~\ref{tab:sigtesting} that the $p$-values for visual inference here are highly variable.

@@

Goodness-of-fit testing for network models is notoriously difficult. Most network models, other than the most simple, lack the necessary asymptotics for developing goodness-of-fit methods \citep{goldenberg09}. Some simulation-based methods have been developed using what \citeauthor{RSienamanual} call \textit{auxiliary statistics} such as the indegree or outdegree distribution on the nodes. In \texttt{RSiena}, the \texttt{sienaGOF} function performs goodness-of-fit testing as follows:
\begin{enumerate}
\item Auxiliary statistics, such as the cumulative outdegree counts on the nodes, are computed on the observed data ($\mathbf{u}_d$) and on $N$ observations simulated from the model ($\mathbf{u}_1 \dots \mathbf{u}_N$). 
\item The mean $\overline{\mathbf{u}}$ and covariance matrix $\mathbf{S}$ are computed from the $N$ simulations, and the Mahalanobis distance, $d_M(\mathbf{u}_d)$ from the observed statistics to the distribution of the simulated statistics is computed:
\begin{equation}\label{eq:mahal}
d_M(\mathbf{u}_d) = \sqrt{(\mathbf{u}_d - \overline{\mathbf{u}})' \mathbf{S}^{-1} (\mathbf{u}_d - \overline{\mathbf{u}})}
\end{equation}
\item The Mahalanobis distance for each of the $N$ simulations is calculated and $d_M(\mathbf{u}_d)$ is compared to this distribution of distances.
\item An empirical $p$-value is found by computing the proportion of simulated distances found in step 4 that are as large or larger than $d_M(\mathbf{u}_d)$.% A model is thus considered a good fit to the data if $p$ is large.
\end{enumerate}
<<gofsiena, fig.height=2.5, fig.cap="An example of what a goodness-of-fit plot from \\texttt{RSiena} looks like. The overlaid boxplots and violin plots show the distribution of each of the outdegree count values on the simulated networks, and the red points and lines are the observed data values.", cache = TRUE>>=
load("data/ansnullpaper.rda")
library(lattice)
#gof1 <- sienaGOF(ansnull, OutdegreeDistribution, varName ="friendship", join = FALSE)
gof1 <- readr::read_rds("data/gof-data.rds")
sims1 <- data.frame(gof1$`Period 1`$Simulations)
dat1 <- gof1$`Period 1`$Observations
dat1 %>% data.frame() %>% gather(outdegree, val) %>%
  mutate(outdegree = parse_number(outdegree)-1) -> dat1
p1 <- sims1 %>%
  mutate(sim = row_number()) %>%
  gather(outdegree, val, X1:X9) %>%
  mutate(outdegree = parse_number(outdegree) - 1) %>%
  filter(outdegree <= 6) %>%
  ggplot(aes(x = outdegree, y = val)) + 
  geom_boxplot(aes(group = outdegree), size = .5, outlier.shape = "x", outlier.size = 5) + 
  geom_violin(aes(group = outdegree), bw = 2, fill = NA) + 
  geom_point(data = dat1, color = 'red') + 
  geom_line(data= dat1, color = 'red') + 
  geom_text(data = dat1, aes(label = val), hjust = -.5) + 
  labs(x = "Outdegree (p = 0.154)", y = "Statistic", 
       title = "Goodness-of-Fit: Outdegree distribution period 1") + 
  ThemeNoNet
p1
@
A plot comparing the data to the simulations is also considered, and a similar plot is shown in Figure~\ref{fig:gofsiena} for the outdegree distribution of small data set, shown in the points and connected lines, with the simulated values of $\mathbf{u}_d$ shown in boxplots and overlaid violin plots.
The \texttt{RSiena} software also provides a Rao score-type test for goodness-of-fit for assessing one or more parameters, the test statistic of which is compared to a Chi-square distribution with $H$ degrees of freedom, where $H$ has the same definition as in Section~\ref{sec:sigtest}. For full detail on the score-type test, see \citet{scoretest}. These methods are both restriced: the \texttt{sienaGOF} method only considers \textit{one} measure on the data and simulations from the model, while the score-type tests only consider \textit{subsets} of parameters, ``nuisance parameters" in \citet{scoretest}, not the entire set of parameters. By using visual inference instead of more traditional statistical methods, we hope to perform a more \textit{holistic} goodness-of-fit test.

@@

<<predictglmm, fig.height=6, fig.cap='Predictions from our GLMM given in Equation~\\ref{eq:glmm}. For new observers of new lineups, the lines show the expected probability of detecting the alternative data in a lineup of size 6 as a function of the parameter value. The proportions detected by our Turk participants for each lineup group are shown by the points, with the probability of picking out the data plot at random shown by a horizontal line at 1/6. The lineup marked as ``outlier" was removed from modeling. Estimated parameter values shown by vertical dotted lines.'>>=
newdata <- read_csv("data/newdata_pred_glmm.csv")
newdata$param <- newdata$test_param
newdata$param <- as.factor(newdata$test_param)
levels(newdata$param) <-  c(1,3,4,2,6,5) 
newdata$param <- as.integer(as.character(newdata$param))
plotdata <- turk22 %>% mutate(type2 = ifelse(sign == 0, sign(initialEst), sign),
                       type2 = ifelse(type == "one", type2, -1)) %>% 
  group_by(pic_id, size, test_param, sign, type2,alt_model) %>% summarize(
    datapick = mean(datapick)
)
plotdata$param <- plotdata$test_param
plotdata$param <- as.factor(plotdata$param)
levels(plotdata$param) <- c(1,3,4,2,6,5) 
plotdata$param <- as.integer(as.character(plotdata$param))
labdat <- data_frame(test_param = "jttp", param = 3, type2 = -1, x = -5, y = .9, label = "outlier")
betahat <- tibble(param = 1:6, size = c(-4.903, 4.893, -3.45, 3.34, 10.091, 1.329))

ggplot() + 
  geom_vline(data = betahat, aes(xintercept = size), linetype = "dotted", color = "grey60") + 
  geom_text(data = betahat, aes(x = size, y = .9, label = paste0("hat(beta)[", param, "]")), parse = T, hjust = 0, color = "grey60") + 
  geom_line(data = newdata, aes(x = size, y = predictfinal, color = as.factor(type2))) + 
  geom_point(data = plotdata, aes(x = size, y = datapick, color = as.factor(type2))) + 
  #geom_text(data = plotdata, aes(x = size, y = datapick, color = as.factor(type2), label = pic_id)) +
  geom_text(data = labdat, aes(x = x, y=y, label = label, color = as.factor(type2)),
            show.legend = F, vjust = 2) + 
  geom_hline(yintercept = 1/6, linetype='dashed') +
  scale_color_brewer(palette = "Dark2", name = "Lineup Type") + 
  facet_wrap(~param, scales = 'free_x', 
             labeller = label_bquote(beta [.(param)])) + 
  labs(x = "Parameter value", y = "Expected probability of detection in a lineup of size 6") + 
  ThemeNoNet + 
  theme(legend.position = 'bottom')

@@

These results do not come without limitations. In VI, the null plots are supposed to play the role of good representatives of the null model. Here, the number of null plots is reduced to five, which has lead to very different conclusions for the same lineup scenario. Furthermore, these results do not generalize to all CTMC models or to one particular subset of CTMC models. The lineups shown are made for only one set of data, and it is not clear that the visual power results will transfer to other situations with different number of actors, different edge densities, or different layout algorithm of the node-link diagram. We can make some generalizations about what participants are picking up on in the lineups based on their feedback and previous research, but we cannot apply our hierarchical model directly to lineups constructed for new data or new models or parameters.

@@

We these methods will be applied to different types of network data and different types of network models. But given the limitations of the node-link visualization, the cognitive load of looking at a lineup is very high for the average observer. More research is needed to apply these methods to larger datasets, different layout algorithms, and different ways of visualizing network data, such as adjacency matrix visualizations, using visual inference to see if similar findings emerge.
